---
title: "Are we more civilized today?"
date: "created 4. oktober 2021, updated `r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
author: "Jens Halfdan Nielsen"



---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE)
```
#  Explore Homicide rates in NW Europe
<!-- In this exercise, you will load a Manuel Eisner's dataset that captures the developments in homicides from Middle Ages to 20th century. -->

**Exploring data and creating visualisations** 

<!-- Upload to Github this .Rmd file with solutions and its .html knitted version.**  -->
<!-- Submit to Brightspace a document that follows the template and contains (1) URL of this script and its knitted version in Github, (2) one plot of your choice,  and (3) answer of 250 words to the Final Question.** -->



This exercise is based on the dataset provided by [OurWorldInData project](https://ourworldindata.org/homicides) based at the Oxford University.

## Long-term trend for Homicides in Western Europe

Understanding how homicide rates have changed prior to the modern era requires the help of historians and archivists. Manuel Eisner, a criminology professor at the University of Cambridge, and his colleagues published the Historical Violence Database \href{https://cjrc.osu.edu/research/interdisciplinary/hvd}: a compilation of data on long-term trends in homicide rates, in addition to qualitative information such as the cause of death, perpetrator and victim. This database is limited to countries with relatively complete historical records on violence and crime – mainly Western Europe and the US.

Starting in the second half of the nineteenth century, these European regions have consistent police records of those accused of murder or manslaughter and annual counts of homicide victims. To go back further in time, reaching as far back as the thirteenth century, Eisner collected estimates (from historical records of coroner reports, court trials, and the police) of homicide rates made in over ninety publications by scholars.

Homicide rates – measured as the number of homicides per 100,000 individuals – up to 1990 are sourced from Eisner’s (2003) publication and the Historical Violence Database. 

Are homicide rates in Europe today lower or higher than in the past? Using the provided dataset, display and describe the long-run homicide rates for the five European regions: Italy, England, Germany, Netherlands and Scandinavia.

```{r library}
# load tidyverse
library(tidyverse)
```

## Loading data from ourworldindata.org

<!--You should always interrogate the source of your data. Who compiled it, from where, what is missing, how representative the data are? Check the data/Metadata.txt to learn about the data provenance. -->

```{r load-data}
#load data for homicide rates in Western Europe
Western_Europe <- read_csv("data/homicide-rates-across-western-europe.csv")
```

## Inspecting the data

<!--How clean and analysis-ready is the dataset? Do you understand what the column names represent? What is the difference between rate and homicide number?-->

```{r inspect}
#inspect element to determine start and end year of data
head(Western_Europe)
tail(Western_Europe)
```

<!--Ok, the data look good except for the column `Homicide rate in Europe over long-term (per 100,000)` which is not very easy to work with.

- Use the `names()` function and assignment key to relabel this column to `homicides_per_100k` -->


```{r relabel column}
# names() function produces the names of columns as a character vector with 4 elements
names(Western_Europe)

# If we can select the fourth element, we can also rename it / reassign a name to it
names(Western_Europe)[4] <- "Homicides per 100k"

# check the new name!
names(Western_Europe)
```

<!--Now, that you have looked at what the data looks like and what it represents, and streamlined it, let's see what big picture it contains.-->

## Overall trends in homicides for Europe: 1300-2000

<!--- use `ggplot()` function and remember the`+` at the end of the line
- chose a meaningful `geom_......()`  for geometry (hint: points are not great)
- load `Year` on the `x` axis and `homicides_per_100k` column in y axis
- to color individual country entries consistently, assign the country column to the argument `color`.
- provide meaningful title and axis labels
- remember to change the `eval` flag so that the code chunk renders when knitted -->

```{r plot-homicides, eval = TRUE}
# create plot for homicide rates in Western Europe in single graph

Western_Europe %>% 
  ggplot(aes(x = Year, y = `Homicides per 100k`, color = Entity))+
  geom_line() +
  theme_minimal()+
  theme(text=element_text(size=18))
  #....CONTINUE WITH YOUR OWN CODE

```
<!--Alright, the homicide rates should all be descending over time. What a comfort. But the viz is not super clear. Let's check the rates for individual countries.-->

## Homicide rates by country

<!--You can visualize each country's trend separately by adding an extra argument to the ggplot, the `facet_wrap()` and feeding it the country column. If in doubt, check your ggplot tutorial and your country column name for exact usage.

- reuse the ggplot from the chunk above
- insert `facet_wrap()` after the specification of geometry to split countries in separate charts
- change the facet "layout" to two columns and three rows so that the trends are easier to see in horizontal layout. -->


```{r eval = TRUE}
# create plots for homicide rates in Western Europe for each country, and clean the graphs for readability!
Western_Europe %>% 
  rename(Country = Entity) %>% 
  ggplot(aes(x = Year, y = `Homicides per 100k`, color = Country))+
  geom_line() +
  facet_wrap(~Country, scales = "free_x", ncol = 2, nrow = 10, strip.position = "top")+
  theme_minimal()
 


```

## Comparing trends: European homicide rates vs average reign of Danish rulers

<!--- Load your Danish king dataset. Hopefully it is tidy and your years and duration of reign are all numeric.
- You need to have a consistent way of plotting the rulers' reign on the x axis, so I recommend you create a midyear column by calculating the middle of each monarch's rule (Hint: `midyear` = `endyear` - (`endyear`-`startyear`)/2)
- Start a ggplot plotting midyear on x axis and duration on y axis
- Try `geom_smooth()` for geometry 
- Provide meaningful labels and a title-->
- Question: How can we characterize the trend compared to the homicides above?

- Answer: As overall the rates of homicides in scandinavia goes down from 1300, the average rule duration for kings goes up. This corresponds well, if we consider that a cause of death for monarchs could be by homicide.

```{r eval = TRUE}
# create plot for average rule duration for Danish monarchs

kings <- read_csv2("data/danish_monarchs.csv")

kings <- kings %>% 
  mutate(duration=end_of_reign - start_of_reign) %>% 
  mutate(midyear = end_of_reign - duration/2) # create midyear vector as ½ of duration (of rule)

glimpse(kings)

kings %>% 
  ggplot(aes(x = midyear, y = duration)) +
  geom_smooth() +
  scale_x_continuous(
    limits = c(900, 2025),
    breaks = seq(900, 2025, by = 100)) + #defines x-axis as period 900-2025, in intervals of 100 yrs 
  theme_minimal() +
  labs(title = "Danish monarchs rule durations",
       x = "Period of rule",
       y = "Duration of rule")
 
```


## **Final tasks:**

<!--1) Plot: In the facetted plot above, move the legend from the current position on the side to below the facets, and label it "Country" instead of "Entity".

2) Rmarkdown: 

  - edit the author of the document, and convert 'Final Tasks' into heading #2 (like the other headings) 
  - add a `floating table of contents` to your Rmarkdown document, 
  - provide informative `chunk-names` and edit flags in your R chunks to `TRUE`, and
  - automatically generate a `timestamp` to show when the document was last updated. (Hint: check the Rmarkdown episode in our Data Carpentry tutorial)-->

###  Final question: 

 **are we more civilized today?** 
 
 In short; yes. If the term 'civilized' is measured as a factor of how many homicides are registered in police reports and court cases, then a clear cut trend can be observed in the graphs above. Murder rates are on a clearly decending trend since the earliest records used, with few minor exceptions/bumps along the way, such as for Italy around 1800 and in Scandinavia shortly before and after the year 1600 (corresponding in time with major national crisis and upheaval from the extensive wars with Sweden and the Holy Roman Empire in the period). The number of homicides today is universally lower than in the past 700 years of observations. Furthermore we can put the homicide rates in perspective to our own calculation of the average rule durations of Danish monarchs. This graph shows an over-all upward trend in rule times over the period, which, at the very least, does not contrast with the lower murder rates. It must be mentioned that the cause of death for Danish kings after 1286 is never directly by homicide, although some are killed in wartime, so the comparison is quite arbitrary! However, the shown trend does not directly contradict the hypothesis that we are indeed more civilized today than in the past, from a homicide perspective.
